Did WW2 Produce an Equivalent to a Modern Computer?

### 2017-06-15 Draft 2/Appendices.tex
Okay so when I say done this time I mean REALLY done

\section{Explanation of Example Turing Machine}\label{appendix:turingexampleexp}

@@

Producing the sequence \texttt{010110111011110111110...}, this machine operates by first printing the symbols \texttt{e e 0} at the start of the tape, before printing the first \texttt{0} of the sequence on the fifth cell, finishing on the third cell and moving to state $\mathbb{A}$.

>>>>

\section{Turing Machine Demonstrating Value \& Marker Squares}\label{appendix:turingexampleexp}

@@

Producing the sequence \texttt{010110111011110111110...} on primary squares, this machine operates by first printing the symbols \texttt{e e 0} at the start of the tape, before printing the first \texttt{0} of the sequence on the fifth cell, finishing on the third cell and moving to state $\mathbb{A}$.

### 2017-06-15 Draft 2/Master.tex
Okay so when I say done this time I mean REALLY done

\title{Did WW2 Produce an Equivalent to a Modern Computer?}
\date{February 23, 2017}
\author{Max Penrose}

@@

\begin{abstract}
Computability, or the potential for a device to solve any solvable problem given to it, is an informal notion. Its definition can be refined with use of the $\lambda$-Calculus or Turing machines, theoretical computational devices which while having simple definitions have functionality similar to that of a computer. Hence, a computer can be defined as a machine which emulates either of these devices fully. It will then be shown that the Z3 computer built by Konrad Zuse implements computability with some success.
\end{abstract}

\section*{Introduction}
Computation rules the modern world - from the internet to the rise of automated labour, today's humanity puts colossal value in the hands of the remarkable machines that are computers. However, the fundamental ideas of computation and calculation lying alongside the transistors and silicon at the heart of this revolution are no less remarkable in themselves, primarily down to their simplicity. Furthermore, few modern computer users understand remotely how basic electronics can operate the complex tasks a computer can perform. Therefore, I feel the natural way to shed light on this question is to look at the earliest and simplest computers, what they could do and what was distinctive about them.

The most significant problem in answering this question is the actual definition of `computer'. In the early 20th Century, prior to developments in this field, any human which evaluated statements through computation was known, logically, as a `computer'. As such, we will explore the qualities that an entity - human, machine or other - must have for it to be able to perform computation.

The 18th Century German polymanth Gottfreid Leibniz proposed the ideal of a `Characteristica Universalis', a universal language in which any possible problem can be expressed, and a decision method which could evaluate any problem in this language \cite{barendregt1984introduction}. This idea neatly defines what a computer is - a machine which when given a statement in this universal language is able to evaluate it\footnote{In fact, it can be shown that there are problems in a universal language which cannot be solved. This was proven indipendently by Alan Turing \cite{turing1936computablenumbers} \& Alonzo Church in 1936, and is known as the Entscheidungsproblem.}. The ideal of producing a simple system which can evaluate any `effectively computable' statment is the main challenge of computation.
% * <mxpen26@gmail.com> 09:38:23 07 Jun 2017 UTC+0100:
% Footnote to Godel's Incompleteness theorum?

Therefore, in order for a machine to be called a `computer', it must be able to evaluate any member of a set of `computable functions' (problems expressed in the universal tongue). However, this is an abstract notion - to clarify it, a simple language in which we can express computable functions must be defined, and a simple method to evaluate them must be found. The two most common methods, $\lambda$-Calculus and the Turing Machine, shall be explored and evaluated in terms of their similarities and differences, before applying this understanding to the early computers.

@@

\section*{The Limits of Computation}

>>>>

\title{Did WW2 Produce an Equivalent to a Modern Computer?}
\author{Max Penrose}

@@

\begin{abstract}
Computability, or the potential for a device to solve any solvable problem given to it, is an informal notion. Its definition can be refined with use of the $\lambda$-Calculus or Turing machines, theoretical computational devices which while having simple definitions produce functionality similar to that of a computer. Hence, a computer can be defined as a machine which emulates either of these devices fully. It will then be shown that the Z3 computer built by Konrad Zuse implements computability with some success.
\end{abstract}

\section*{Introduction}
Computation rules the modern world - from the internet to the rise of automated labour, today's humanity puts colossal value in the hands of these remarkable machines. However, the fundamental ideas of computation and calculation lying alongside the transistors and silicon at the heart of this revolution are no less remarkable in themselves, primarily down to their simplicity. Furthermore, few modern computer users understand remotely how basic electronics can operate the complex tasks a computer can perform. Therefore, I feel the natural way to shed light on this question is to look at the earliest and simplest computers, what they could do and what was distinctive about them.

The most significant problem in answering this question is the actual definition of `computer'. In the early 20th Century, prior to developments in this field, any human which evaluated statements through computation was known, as a `computer'. We will therefore explore the qualities that an entity - human, machine or other - must have for it to be able to perform computation.

The 18th Century German polymanth Gottfreid Leibniz proposed the ideal of a `Characteristica Universalis', a universal language in which any possible problem can be expressed, and a decision method which could evaluate any problem in this language \cite{barendregt1984introduction}. This idea is a final aim for computation - a system which when given a statement in this universal language is able to evaluate it. The ideal of producing a simple system which can evaluate any `effectively computable' statment is the main challenge of computation.

Therefore, in order for a machine to be called a `computer', it must be able to evaluate any member of a set of `computable functions' (problems expressed in the universal tongue). However, this is an abstract notion - to clarify it, a simple language in which we can express computable functions must be defined, and a simple method to evaluate them must be found. The two most common methods, $\lambda$-Calculus and the Turing Machine, shall be explored and evaluated in terms of their similarities and differences, before they are used to evaluate early computers.

@@

\section*{Epilogue: Limitations of Computation}

### 2017-06-15 Draft 2/Section 2 - Turing Machines.tex
Okay so when I say done this time I mean REALLY done

An example of a more complex Turing machine is shown below, which generates the sequence \texttt{010110111011110111110...} again on alternate cells, beginning at the fifth cell. \cite{turing1936computablenumbers}

\medskip\noindent\begin{tabu} to \textwidth{XXXX}
    \multicolumn{2}{c}{\textit{Configuration}} & \multicolumn{2}{c}{\textit{Behavior}} \\
    \textit{state} & \textit{symbol} & \textit{transition function} & \textit{final state} \\
    \hhline{====}
    \multirow{1}{*}{$\mathbb{Q}$} & Any        & P\texttt{e}, R P\texttt{e}, R,  P\texttt{0}, R, R, P\texttt{0}, L, L & $\mathbb{A}$ \\
    \hhline{----}
    \multirow{2}{*}{$\mathbb{A}$} & \texttt{0} & R, P\texttt{x}, L, L, L                   & $\mathbb{A}$ \\
                                  & \texttt{1} &                                  & $\mathbb{B}$ \\
    \hhline{----}
    \multirow{2}{*}{$\mathbb{B}$} & Any        & R, R                             & $\mathbb{B}$ \\
                                  & None       & P\texttt{1}, L                            & $\mathbb{C}$ \\
    \hhline{----}
    \multirow{3}{*}{$\mathbb{C}$} & \texttt{x} & E, R                             & $\mathbb{B}$ \\
                                  & \texttt{e} & R                                & $\mathbb{D}$ \\
                                  & None       & L, L                             & $\mathbb{C}$ \\
    \hhline{----}
    \multirow{2}{*}{$\mathbb{D}$} & Any        & R, R                             & $\mathbb{D}$ \\
                                  & None       & P\texttt{0}, L, L                         & $\mathbb{A}$ \\
\end{tabu}

\noindent $\Rightarrow \mathbb{Q}$

The operation of such a machine is significantly more complex than the earlier one - a full explanation of its operation can be found in appendix \ref{appendix:turingexampleexp}. Key to the operation is the fact that the tape begins with the sequence \texttt{e e 0} at the start of the tape (this method shall be used in future), and that cells are seperated into value cells and marker cells. On value cells (odd-numbered cells) the symbols of the output sequence are printed. Only the symbols \texttt{0} and \texttt{1} will ever be printed on these cells. To the right of each value cell is a label cell (an even-numbered cell). These cells will either be left empty or will contain a 'marker' symbol, which labels the symbol on the value cell to the left. These symbols can be erased and overwritten, and are not part of the output sequence. Only one label cell is needed to contain multiple markers, or more descriptive markers for a single value cell, as this can be achieved with a sufficiently rich variety of marker symbols. This method is frequently used by Alan Turing in \cite{turing1936computablenumbers}.

It is clear from this example that the operation of the Turing Machine is very different from that of an ordinary computer. However, the core components are still mostly present - conditionals in the form of the symbol tests in the second column, variables in the form of labels, and continuity and looping when a state triggers itself - a form of recursion. The end result is the same - a solution is produced in the form of the computable sequence written on the value squares, and the machine can continue this sequence without halting indefinitely. Unfortunately, the operation of this machine is somewhat complex and obfuscated. To simplify this, we will introduce a type of function for the Turing machine.

@@

The use of \textit{m}-functions greatly simplifies the table descriptions, as certain operations are common to many machines. Of course, any table description that includes or refers to \textit{m}-functions can be written as an ordinary table as described in \cref{sec:TMacIntro}. This is possible because of the finite nature of both states and symbols; simply by listing in separate rows under the same state every permutation of the finite symbols and states that the function could possibly take substituted in.

@@

In this section a set of common and useful \textit{m}-functions are defined that will be used regularly in future.

@@

At this point, the similarity between Turing machines and modern computers becomes very clear. The machine is able to store information, from integers to strings of characters to images, as binary sequences of value square \texttt{0}s and \texttt{1}s - these shall be stored on the tape as memory. The first data structure we shall define is the simple boolean - only 1 bit (1 value square) long. In order to identify these sequences on the tape as booleans, we shall label the first character of the boolean with an identifier (we shall call these pointers) unique to that boolean variable - for example the symbol \texttt{\textbf{B}myvar} (bear in mind this is one symbol, not six) would mark a boolean named \textit{myvar}. Let us define some \textit{m}-functions to define and assign values to variables of the boolean type:

@@

Most remarkably, Turing machines are in fact able to also emulate $\lambda$-Caluculus. This shall not be demonstrated directly, but it seems intuitive that a Turing machine could carry out processes of abstraction and application since they are at heart simple algorithmic processes; a Turing macihne would evaluate $\lambda$-terms in a similar way to how the human mind would. This leads us onto the remarkable fact that since Turing machines and $\lambda$-Calculus can produce Universal Machines for each other, each has the potential to perform any task the other can perform, and hence they are equal in power. Furthermore, many other theoretical methods of computation such as Kurt G\"odel's General Recursive Functions also share this equivalence \cite{copeland2002ctt}.

>>>>

To aid the operation of Turing machines, the tape shall begin with the sequence \texttt{e e 0}, and cells shall be seperated into value cells and marker cells. On value cells (odd-numbered cells) the symbols of the output sequence are printed. Only the symbols \texttt{0} and \texttt{1} will ever be printed on these cells. To the right of each value cell is a label cell. These cells will either be left empty or will contain a 'marker' symbol, which labels the symbol on the value cell to the left; these are not part of the output sequence. This method is frequently used by Alan Turing in \cite{turing1936computablenumbers}, and an example machine can be found in appendix \ref{appendix:turingexampleexp}.

@@

The use of \textit{m}-functions greatly simplifies the table descriptions, as certain operations are common to many machines. Of course, any table description that includes or refers to \textit{m}-functions can be written as an ordinary table as described in \cref{sec:TMacIntro}. This is possible because of the finite nature of both states and symbols.

@@

@@

At this point, the similarity between Turing machines and modern computers becomes very clear. The machine is able to store information, from integers to strings of characters to images, as binary sequences of value square \texttt{0}s and \texttt{1}s. The first data structure we shall define is the simple boolean - only 1 bit (1 value square) long. In order to identify these sequences on the tape as booleans, we shall label the first character of the boolean with an identifier (we shall call these pointers) unique to that boolean variable - for example the symbol \texttt{\textbf{B}myvar} (bear in mind this is one symbol, not six) would mark a boolean named \textit{myvar}. Let us define some \textit{m}-functions to define and assign values to variables of the boolean type:

@@

Most remarkably, Turing machines are in fact able to also emulate $\lambda$-Caluculus. This shall not be demonstrated directly, but it seems intuitive that a Turing machine could carry out processes of abstraction and application since they are at heart simple algorithmic processes; a Turing machine would evaluate $\lambda$-terms in a similar way to how the human mind would. This leads us onto the remarkable fact that since Turing machines and $\lambda$-Calculus can produce Universal Machines for each other, each has the potential to perform any task the other can perform, and hence they are equal in power. Furthermore, many other theoretical methods of computation such as Kurt G\"odel's General Recursive Functions also share this equivalence \cite{copeland2002ctt}.

### 2017-06-15 Draft 2/Section 4 - Zuse's Z3.tex
Okay so when I say done this time I mean REALLY done

At this point, we have a complete definition of `computer' - a device which is able to fully implement the concept of effective computability. This can be done most simply through the emulation of an effectively computable method of computation, such as a Turing Machine. In order to address the question of whether this was achieved during WW1 we must apply this test to the physical instruments created during this period. The

Z3, completed in Berlin in 1941 by Konrad Zuse, was arguably the first programmable computer ever built. A far faster and more reliable version of its two predecessors, Z1 and Z2, Z3 was advanced because of its ability to carry out long sequences of calculations, written as instructions on punched tape, with the potential to store values in memory. While still very simplistic, and not directly intended for universal computation, Z3 is structurally far closer to achieving this aim than other machines of the time; Hence, in this section I shall attempt to implement a Universal machine in Z3 in order to determine its nature as a computing machine.

@@

\subsubsection{Data Storage}

\textit{Basically it can store decimals, and these integers are referred to by a numeric 'address'. for the moment we'll use letters instead of numbers because we're cool like that}

>>>>

In order to address the question of whether this was achieved during WW1 we must explore how well this concept was implemented in the machines of this period. Z3, completed in Berlin in 1941 by Konrad Zuse, was arguably the first programmable computer ever built. A far more reliable version of its predecessors, Z1 and Z2, Z3 was advanced because of its ability to carry out long sequences of calculations written as instructions on punched tape, on top of its ability to store values in memory. While still very simplistic, and not directly intended for universal computation, Z3 was structurally far closer to achieving this aim than other machines of the time. In this section I shall therefore attempt to implement a Universal machine in Z3 in order to determine its nature as a computing machine.

@@

Z3 was able to store floating-point decimals in its memory\footnote{Each 22-bit floating-point number was stored through the use of metal rods which could be in one of two states to represent bits. In  something similar to standard form, each number was composed of an `exponent' representing the order of magnitude of the number, and a `significand' that represented the digits of the number; see \cite{rojas1997z3architecture} for details.}, which could be referred to by the programmer through the use of a numeric address of a particular data slot. For the sake of clarity, we shall use single lowercase letters to represent addresses rather than the 6-bit unisgned integers of Z3.
